An assessment of the biodiversity - ecosystem function relationship in southern African woodlands

### 2020-04-07 manuscript/regional_befr_manuscript.tex
Added some comments

% Define page geometry
\usepackage{geometry}
\geometry{left=2.2cm,
	right=2.2cm,
	top=2.2cm,
	bottom=2cm}
\parskip 0.15cm
\setlength{\parindent}{0cm}
\usepackage{pdflscape}

@@

@@

\section*{Biosketch}
\newpage{}

@@

\normalsize{Running title: Woodland functionality in southern Africa}

@@

\textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship in highly disturbed and ecophysiologically stressful systems is less clear. Here, we explore the relationship between tree biodiversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density and resource availability, to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.

@@

\textbf{Methods:} We used a network of \nplots{} savanna and woodland plots located across the southern African sub-continent in which each tree >5 cm diameter was measured. We used Structural Equation Modelling and path analysis to determine the relationship between tree species diversity and aboveground biomass, with comparison to the effects of resource availability and along a gradient of stem density.

\textbf{Results:} A positive effect of tree species diversity on biomass was demonstrated, observed largely as an indirect effect of increasing woodland structural diversity. We also found that the effect of tree species diversity on biomass increases with stem density. Finally, we found that resource availability affects biomass in southern African woodlands largely indirectly via its effect on species diversity.

\textbf{Main conclusions:} Tree species diversity plays an important role in determining the ecosystem functionality and structure of southern African woodlands. Our results demonstrate the importance of considering environmental conditions and woodland structural attributes to more accurately describe the relationship between biodiversity and ecosystem functionality.

@@

Numerous studies have shown relationships between biodiversity and ecosystem function (e.g. \citealt{Liang2016, Cardinale2009}). The strength and direction of these observed Biodiversity - Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}. Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}. These studies support early findings from small scale experimental studies predominantly in grassland patches, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.

Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the degree of biotic activity within an ecosystem \citep{Jax2005}. This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass. The frequently reported and intuitive relationship between biodiversity and ecosystem functionality invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others' functional contribution.

Compared to other forested ecosystems, dry tropical woodlands are highly structured by disturbance, mainly through fire, and in Africa notably by herbivory also \citep{}. High levels of disturbance may weaken the role of competition in determinging local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as apparent in dry woodland ecosystems. Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{Lasky2014, Tobner2016}. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth. In European forests \citep{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggest that in dry ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function. This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation.

% COME BACK TO THIS: In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}. It is important therefore to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on woody biomass in analyses at large spatial scales. \citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}. Similarly, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in a wet Panamanian tropical forest.

The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}. A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEFR mechanism described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEFR and it's environmental determinants.

Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (\autoref{clust_map}). The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}. Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species. Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}. Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely un-quantified consequences for ecosystem structure and functionality.

A small number of studies in southern African savannas and woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. However, studied conducted over small environmental gradients often find that at local scales, diversity shows a strong effect on ecosystem function, but at large scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Small scale studies may therefore be over-estimating the effect of diversity. Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass \citep{}, it is important to sample across geographic and environmental gradients to be able to infer a truly regional scale relationship between diversity and biomass.

@@

\citet{Solbrig1996} writes that southern African woodlands possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species. This structural diversity may be one mechanism through which diversity influences woody biomass. \citep{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest. They also found that trees growing in species rich neighbourhoods exhibited enhanced biomass production. Occupation of multiple canopy layers allows a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity. This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas, which are structured by disturbance rather than competition. Instead, disturbance history may override the effects of species diversity on structural diversity nullifying the effects of species diversity on structural diversity.

% High levels of disturbance in southern African woodlands may moderate the observable BEFR through it's effect on ecosystem composition. Fire disturbance in forests has been linked to abundance dependent mortality among smaller tree stems \citep{Roques2001, Staver2009, Bond2005}. A dominant and competitive species which tends to produce many small fast-growing stems is more likely to experience mortality during a fire as the higher surface area to volume ratio of the trunk increases the likelihood of combustion. These fast-growing bushy species often proliferate in southern African woodlands when fire is deliberately excluded, leading to the growth of thicket vegetation which precludes the growth of high biomass tree species \citep{Higgins2007}. Additionally, the dominance of a single species is likely to decrease overall functioning by decreasing niche space utilisation \citep{Cardinale2002}.
% KGD: This text is useful in demonstrating your biological understanding of the ecosystem. I think it could be useful to include some of this biological knowledge in the paragraph above that ends with the statements on disturbance (the paragraph starting with Slobrig 1996). Or you could include this whole paragraph just after that one. The co-authors will reign us in if this kind of logic is not sensible for the ecosystems we are studying.

In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship across southern African savannas and woodlands, using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}). We compared the relative effects of tree species biodiversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility. We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type based on major tree species compositional units as a factor in our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality and assess the generality of our results. We used Structural Equation Modelling (SEM) and path analysis as a preferred method to simultaneously account for environmental and biotic factors, which may interact their effect on ecosystem structure and therefore biomass. Initially, we made three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect size of species diversity on woody biomass will increase with stem density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.

@@

The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in so-called miombo woodlands (\autoref{clust_map}, \citealt{White1983}). The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. The study sites span the core climate space of the region as a whole. The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland area as defined by \citet{White1983}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.

Plots were chosen from a larger pool of 5395 plots based on the quality and completeness of data collection, and plot setup. Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool. Only plots >0.1 hectares were used in analysis, as area based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates. Only plots with a stem density >10 stems ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.

Many plots provided by the Zambian Forestry Commission were arranged in clusters of up to four 20x50 m plots, 20 metres apart. Plots within each cluster were aggregated before the plot dataset filtering described above and treated as a single plot in analyses.

@@

Climatic data were collected from the ECMWF ERA5 dataset, generated using Copernicus Climate Change Service Information \citep{ERA5}. Values of Mean Annual Temperature (MAT) and Mean Annual Precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot. Temperature seasonality (TS) and precipitation seasonality (PS) were both calculated as the mean of the coefficient of variation of daily MAT and MAP, respectively, for each of the 18 years of available data. Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC), percentage soil organic carbon by volume (Org. C \%), and percentage soil sand content by volume (Sand \%). These data are a modelled product derived from various remotely sensed and directly measured data sources.

@@

To measure tree species abundance evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} (\autoref{shannon_equit}) was calculated:

\begin{equation}
	\begin{gathered}
		E_{H'} = \frac{H'_{e}}{\ln{S}} \\
	\end{gathered}
	\label{shannon_equit}
\end{equation}

Where $H'_{e}$ is an estimation of the Shannon diversity index of trees by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the extrapolated tree species richness in the plot, using the \verb|ChaoRichness()| function. We calculated tree structural diversity for each plot by calculating the coefficient of variation of DBH (DBH CV) and tree height (Height CV).

@@

Plots were assigned to vegetation type groups based on tree species composition. Groups were identified in \citet{Fayolle2018} in an Africa wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data. Groups were identified using unconstrained correspondence analysis and ordination, followed by clustering based on dominant ordination axes. Plot data used in this study occurred in four vegetation type groups. See \autoref{clust_summ} for a description of each vegetation cluster and \autoref{clust_map} for the spatial distribution of plots from each of these clusters .

@@

Structural Equation Models (SEM) investigated the determinants of AGB. All SEMs were constructed and analysed in the \verb|lavaan| package \citep{lavaan} in R version 3.6.0 \citep{R2019}. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}. A key aspect to our decision to use SEMs is that they can explicitly model and partition variance to indirect effects, which is challenging in standard multiple regression. Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable measures in our dataset. Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.

@@

The factor loadings of the observed variable assumed to contribute most to each latent variable were set to 1 as per convention, with other observed variables being allowed to vary \citep{Beaujean2014}.  We tested the robustness of our assumptions with a chi-squared test of all possible combinations of observed variable factor loadings set to 1, while ensuring no factor loadings were in excess of 1. We found no significant difference between model specifications. Full Information Max-Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Cham2017}.

>>>>

% Define page geometry
\usepackage{geometry}
\geometry{left=2.2cm,
	right=2.2cm,
	top=2.2cm,
	bottom=2cm}
\parskip 0.15cm
\setlength{\parindent}{0cm}
\usepackage{pdflscape}
\usepackage[document]{ragged2e}

@@

\usepackage[revision]{revdiff}

@@

This work is funded by a NERC E3 Doctoral Training Partnership PhD studentship at the University of Edinburgh (J. L. Godlee, Grant No. NE/L002558/1). The data provided for this study was constributed by a number of independently funded projects and was assembled and prepared by SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands), an activity of the Miombo Network and NERC-funded project (Grant No. NE/P008755/1). We thank all data providers and the field assistance they received when collecting plot data.

\section*{Biosketch}

SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands, \url{https://seosaw.github.io}) aims to understand the response of southern African woodlands to global change. The goal of SEOSAW is to produce novel analyses of the determinants of ecosystem structure and function for the southern Africa region, based on syntheses of plot data. Additionally the group hopes to develop infrastructure for a long-term regional plan for plot remeasurement in the southern African region. While working on a multitude of diverse projects in the dry tropics at large, all authors have a broad interest in community ecology and ecosystem assemblage in southern African woodlands.

\newpage{}

@@

\normalsize{Running title: Ecosystem function in southern African woodlands}

@@

\textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance and ecophysiological stress, is less clear}. Here, we explore the relationship between tree biodiversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density, resource availability \rnew{and disturbance through fire}, to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.

@@

\textbf{Methods:} We used a network of \nplots{} savanna/woodland plots located across the southern African sub-continent in which each tree >10 cm diameter was measured \rnew{and identified to species level}. We used structural equation modelling and path analysis to determine the relationship between tree species diversity and aboveground woody biomass, accounting for the interactive effects of resource availability and along a gradient of stem density.

\textbf{Results:} A positive effect of tree species diversity on aboveground biomass was demonstrated, observed largely as an indirect effect of increasing woodland structural diversity. We also found that the effect of tree species diversity on biomass increases with stem density. Finally, we found that resource availability affects biomass in southern African woodlands largely indirectly via its effect on species diversity.

\textbf{Main conclusions:} The study underlines the close association between tree diversity, ecosystem structure and function of highly disturbed southern African savannas and woodlands. Our results demonstrate the importance of including environmental conditions and vegetation type in models to accurately describe a general relationship between biodiversity and ecosystem function at a regional level. Biodiversity loss predominantly by human actions in southern Africa may have detrimental outcomes for ecosystem function, particularly in species poor Baikiaea woodlands, which showed the strongest biodiversity - ecosystem function relationship.

@@

Scientific interest in the relationship between species richness and ecosystem function springs from both an interest in the factors which structure ecological communities \citep{}, and a more applied interest in determining the effect of global biodiversity loss on ecosystem form and function \citep{}. Numerous studies have shown relationships between biodiversity and ecosystem function (e.g. \citealt{Liang2016, Hooper2012, Cardinale2009}). The strength and direction of these observed Biodiversity-Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016}. Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}. These studies support early findings from small scale experimental studies predominantly in grassland patches, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.

Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the degree of biotic activity within an ecosystem \citep{Jax2005}. This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass. The frequently reported and intuitive relationship between biodiversity and ecosystem function invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others' functional contribution.

Compared to \rchange{other forested}{forest} ecosystems, dry tropical woodlands and savannas are highly structured by disturbance, mainly through fire \rchange{and in Africa notably by herbivory also}{and herbivory, with African savannas possessing large herbivores absent from other savannas} \citep{Sankaran2008, Levick2009}. \rnew{Disturbance via human activities such as timber extraction and charcoal processing is also common in African woodlands, often causing high levels of disturbance in localised areas \citep{}}. High levels of disturbance may weaken the role of competition in determinging local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as apparent in dry woodland/savanna ecosystems. Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces which influence ecosystem function\rnew{s} \citep{Lasky2014, Tobner2016}. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations \rchange{in environments which are more hostile to growth}{under limiting environmental conditions such as low water availability \citep{} or high maximum temperature \citep{}}. Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggest that in dry ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function. This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation in order to characterise a generalisable BEFR.

The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}. A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEFR mechanism described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEFR and its environmental determinants. Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{White1983, Ratnam2011, Ryan2016} (\autoref{clust_map}). The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}. Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species. Southern African savannas and woodlands (SAWs) are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}. Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely un-quantified consequences for ecosystem structure and function.

A small number of studies in SAWs, all of which were restricted in the spatial scope to a small region of miombo woodland, have found that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. The results of these fine scale studies concur with similar studies in other biomes \citep{}. Studies of the BEFR often find that at fine scales, diversity shows a strong effect on ecosystem function, but at broad scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Due to the highly variable environmental conditions within which SAWs occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass \citep{}, it is important to sample across geographic and environmental gradients to gain understanding of the spatial variation in the relationship between diversity and biomass.

@@

In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}. It is important therefore to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on woody biomass in analyses at broad spatial scales. \citet{Sankaran2005} used data from 854 African woodland field sites to show that \rnew{below a threshold of \textapprox{}650 mm MAP}, precipitation sets sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}. Similarly, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in a wet Panamanian tropical forest.

\citet{Solbrig1996} writes that SAWs possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species. This structural diversity may be one mechanism through which diversity influences woody biomass. \citet{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest. They also found that trees growing in species rich neighbourhoods exhibited enhanced biomass production. Occupation of multiple canopy layers allows a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity. This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas, which are structured by disturbance rather than competition. Instead, disturbance history may override the effects of species diversity on structural diversity nullifying the effects of species diversity on structural diversity.

High levels of disturbance in SAWs may moderate the observable BEFR through its effect on ecosystem composition. Fire disturbance in forests has been linked to abundance dependent mortality among smaller tree stems \citep{Roques2001, Staver2009, Bond2005}. Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree. Selection effects may therefore be more important in maximising ecosystem function in disturbance prone woodlands. If the regional species pool contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high AGB under an intense disturbance regime.

In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship across southern African savannas and woodlands (SAWs), using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}). We used aboveground woody biomass of trees and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility. We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type based on major tree species compositional units as a factor in our analyses to understand how species composition as well as species diversity affected ecosystem function and assess the generality of our results. We used Structural Equation Modelling (SEM) and path analysis as a preferred method to simultaneously account for environmental and biotic factors, \rchange{which may interact their effect}{which may have interacting effects} on ecosystem structure and therefore biomass. Initially, we made three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strenghth of the effect} of species diversity on woody biomass will increase with stem density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.

@@

The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the \rchange{so-called miombo woodland eco-region}{miombo ecoregion} (\autoref{clust_map}, \citealt{White1983}). The study \rchange{region spans}{area spacs the core climate space of the region}, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \rold{as defined by} \citep{White1983}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.

Plots were chosen from a larger pool of 5395 plots based on the quality and completeness of data collection, and plot setup. Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands (SAWs) \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool. Only plots >0.1 hectares were used in analysis, as area based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates. Only plots with a stem density >10 stems ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.

Many plots provided by the 2005-2008 Zambian Integrated Land Use Assessment \citep{Mukosha2009} were arranged in clusters of up to four 20x50 m plots, 20 metres apart. Plots within each cluster were aggregated before the plot dataset filtering described above and treated as a single plot in analyses.

@@

Climatic data were collected from the ECMWF ERA5 dataset, generated using Copernicus Climate Change Service Information \citep{ERA5}. Values of \rchange{Mean Annual Temperature (MAT)}{MAT} and \rchange{Mean Annual Precipitation (MAP)}{MAP} were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot. Temperature seasonality (TS) and precipitation seasonality (PS) were both calculated as the mean of the coefficient of variation of daily MAT and MAP, respectively, for each of the 18 years of available data. Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot centre \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC), percentage soil organic carbon by volume (Org. C \%), and percentage soil sand content by volume (Sand \%). These data are a modelled product derived from various remotely sensed and directly measured data sources.

@@

To measure tree species abundance evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} (\autoref{shannon_equit}) was calculated as the ratio of estimated Shannon diversity to the natural log of estimated species richness. Abundance evenness allows for greater niche complementarity at small scales due to an increased spatial heterogeneity of functional traits. We calculated tree structural diversity for each plot by calculating the coefficient of variation of DBH (DBH CV) and tree height (Height CV).

% \begin{equation}
% 	\begin{gathered}
% 		E_{H'} = \frac{H'_{e}}{\ln{S}} \\
% 	\end{gathered}
% 	\label{shannon_equit}
% \end{equation}
% 
% Where $H'_{e}$ is an estimation of the Shannon diversity index of trees by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the extrapolated tree species richness in the plot, using the \verb|ChaoRichness()| function.

@@

Plots were assigned to vegetation type groups based on tree species composition. Groups were defined in \citet{Fayolle2018} in an Africa wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data. Group identification was conduscted using unconstrained correspondence analysis and ordination, followed by clustering based on dominant ordination axes. Plot data used in this study occurred in four vegetation type groups. See \autoref{clust_summ} for a description of each vegetation cluster and \autoref{clust_map} for the spatial distribution of plots from each of these clusters .

@@

Structural Equation Models (SEM) investigated the determinants of AGB. All SEMs were constructed and analysed in the \verb|lavaan| package \citep{lavaan} in R version 3.6.0 \citep{R2019}. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}. A key aspect to our decision to use SEMs is that they can explicitly model and partition variance to indirect effects, which is challenging in standard multiple regressions. Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable measures in our dataset. Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.

@@

The factor loadings of the observed variable assumed to contribute most to each latent variable were set to 1 as per convention, with other observed variables being allowed to vary \citep{Beaujean2014}.  We tested the robustness of our assumptions with a chi-squared test of all possible combinations of observed variable factor loadings set to 1, while ensuring no factor loadings were in excess of 1. We found no significant difference between model specifications (p>0.05). Full Information Max-Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Cham2017}.

We assessed the role of structural diversity and species diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity. To account for variation in stem density which may covary with species diversity we also included it as an predictor in our model. To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation cluster separately. We compared unstandardised path coefficients among these vegetation cluster scale models to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB. Path coefficients show the effect of a path with other paths of inference held constant. Models were estimated using the ``MLM'' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}. Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB. We critically assess model fit in each case, taking into consideration the recommendations of \citet{Hu1999} which define threshold values of acceptability for these model fit indices: CFI >0.85, TLI >0.85, RMSEA <0.15, alongside our judgement of the model estimates.

To explore the hypothesis that complementarity effects increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create \subn{} datasets of similar size with varying median stem density. We used each of these datasets separately to fit the model including only tree species and structural diversity latent variables to predict AGB. We excluded the effect of stem density on AGB and the correlation between stem density and species diversity from this model as we were deliberately controlling stem density in our subsampling. We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stem density of subsampled dataset.

We incorporated environmental covariates into our model to understand the relative effects of moisture availability and soil fertility on AGB both directly and indirectly via species diversity and stem density. We compared standardised path coefficients between paths in the model to understand the relative contribution of each path to explain variance in AGB. Vegetation type specific models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation clusters were narrow in their climate space leading to a lack of variance particularly in moisture availability.
